Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-12 paper/paper.tex
edits in appendices

$C = 0$. From \eqref{eq:c-gauss-entropy-derivation} the entropy of a proper $\cplx$-Gaussian is

@@

@@

We consider two priors for Bayesian dropout: log-uniform and $\cplx$-Gaussian empirical
Bayes, \citep{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}.
For a fully factorized approximation $q(\omega)$ and factorized prior belief $\pi(\omega)$,
the divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q(\omega) \| \pi(\omega))
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = \sum_{ij} \bigl(
        - \mathbb{H}(q(\omega_{ij}))
        - \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})} \log{\pi(\omega_{ij})}
      \bigr)
    \,.
\end{equation}
We use approximation \eqref{eq:c-gauss-vi} and omit subscripts ${ij}$ below for brevity.

@@

\begin{equation}. \label{eq:nonchisq-density}

@@

In \eqref{eq:ard-kl-div-real} we need the expected logarithm of $W$:

@@

    &= \int_0^\infty f_W(x) \log x dx

@@

    \biggr) \log x dx

@@

Note that $\psi$ satisfies: $

@@

Differentiating the series within $g_m$ yields
\begin{equation*}  \label{eq:g-m-deriv}
  \frac{d}{d x} g_m(x)
    % = e^{-x} \sum_{n\geq 1} \frac{x^{n-1}}{(n-1)!} \psi(n + \tfrac{m}2) - g_m(x)
    % = e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} \psi(n + \tfrac{m}2 + 1) - g_m(x)
    % = e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} (
    %     \psi(n + \tfrac{m}2 + 1) - \psi(n + \tfrac{m}2)
    % )
    % = e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} \tfrac1{n + \tfrac{m}2}
    = e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} (n + \tfrac{m}2)^{-1}
    = \cdots
    \,.
\end{equation*}
Since $
    e^t  = \sum_{n\geq 0} \tfrac{t^n}{n!}
$ on $\mathbb{R}$, for $\alpha\neq 0$ we have $
  \int_0^x t^{\alpha-1} dt
    = \tfrac{x^\alpha}{\alpha}
$ and
\begin{equation*}  \label{eq:g-m-deriv-int-1}
  \sum_{n\geq 0} \int_0^x \frac{t^{n+\alpha-1}}{n!} dt
    = \int_0^x \sum_{n\geq 0} \frac{t^{n+\alpha-1}}{n!} dt
    = \int_0^x t^{\alpha-1} e^t dt
    \,,
  % the MCT ([0, x], \mathcal{B}([0, x]), dx)
\end{equation*}
by the monotone convergence theorem, whence
\begin{equation}  \label{eq:g-m-deriv-int-2}
  \cdots
    = x^{-\tfrac{m}2} e^{-x} \sum_{n\geq 0}
        \frac{x^{n + \tfrac{m}2}}{n!} (n + \tfrac{m}2)^{-1}
    % = x^{-\tfrac{m}2} e^{-x} \sum_{n\geq 0}
    %     \frac1{n!} \int_0^x t^{n + \tfrac{m}2 - 1} dt
    % = x^{-\tfrac{m}2} e^{-x}
    %     \int_0^x t^{\tfrac{m}2 - 1} \sum_{n\geq 0} \frac{t^n}{n!} dt
    = x^{-\tfrac{m}2} e^{-x}
        \int_0^x t^{\tfrac{m}2 - 1} e^t dt
    = \cdots
    \,.
\end{equation}
The substitution $u^2 = t$ on $[0, \infty]$ with $2u du = dt$ yields
$$
\frac{d}{d x} g_m(x)
  = x^{-\tfrac{m}2} e^{-x}
      \int_0^x t^{\tfrac{m}2 - 1} e^t dt
  = 2 x^{-\tfrac{m}2} e^{-x}
      \int_0^{\sqrt{x}} u^{m - 1} e^{u^2} du
  \,. $$
In particular, for $m=1$ we get
\begin{equation}  \label{eq:g-m-series-one}
  g_1(x)
    = - \gamma - 2 \log{2}
    + e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} \sum_{p=1}^n \frac2{2p - 1}
    \,,
\end{equation}
and the derivative \eqref{eq:g-m-deriv} can be computed thus
\begin{equation}  \label{eq:g-m-deriv-one}
  \frac{d}{d x} g_1(x)
    = 2 \tfrac{F(\sqrt{x})}{\sqrt{x}}
    \,,
\end{equation}
using $
  F \colon \mathbb{R} \to \mathbb{R}
  \colon x \mapsto e^{-x^2} \int_0^x e^{u^2} du
$ (Dawson's integral) a special function.

The derivative of \eqref{eq:ard-kl-div-real} with respect to $\alpha$ is
$$
\tfrac{d}{d \alpha} g_1(\tfrac1{2\alpha})
  = -\tfrac1{2 \alpha^2} g_1'(\tfrac1{2\alpha})
  % = -\tfrac1{2 \alpha^2} 2 \tfrac{F(\sqrt{\tfrac1{2\alpha}})}{\sqrt{\tfrac1{2\alpha}}}
  % = -\tfrac1{2 \alpha^2} 2 \tfrac{F(\tfrac1{\sqrt{2\alpha}})}{\tfrac1{\sqrt{2\alpha}}}
  = -\tfrac1{\alpha} \sqrt{\tfrac2{\alpha}} F(\tfrac1{\sqrt{2\alpha}})
  \,. $$
However, since $\alpha$ is nonnegative, it is typically parameterized via its logarithm
and computed when needed. Thus the gradient w.r.t $\log \alpha$ is
\begin{equation}  \label{eq:real-kl-div-deriv-log}
  \frac{d}{d\log \alpha}
    \tfrac12 \mathbb{E}_{z \sim \mathcal{N}(0,1)}
    \log \lvert z + \tfrac1{\sqrt{\alpha}} \rvert^2
    = \frac12 \frac{d\alpha}{d\log \alpha}
      \frac{d}{d\alpha} \bigl(\mathbb{E}\cdots \bigr)
    % = - \frac\alpha2 \tfrac1{\alpha}
    %   \sqrt{\tfrac2{\alpha}} F(\tfrac1{\sqrt{2\alpha}})
    = - \tfrac1{\sqrt{2\alpha}} F(\tfrac1{\sqrt{2\alpha}})
    \,.
\end{equation}

We replicated the simulation experiment in \citep{molchanov_variational_2017} and recovered
the same nonlinear approximation result reported therein. We sampled $10^7$ draws of  and 
The coefficients are within
$1\%$ relative tolerance of the reported.
$$
- KL(\mathcal{N}(w \mid \mu, \alpha \mu^2) \|
    \tfrac1{\lvert w \rvert})
  % = \tfrac12 \log \alpha
  %   - \mathbb{E}_{\xi \sim \mathcal{N}(1, \alpha)}
  %   \log{\lvert \xi \rvert} + C
  \approx
    k_1 \sigma(k_2 + k_3 \log \alpha) + C \big\vert_{C = -k1}
    - \tfrac12 \log (1 + e^{-\log \alpha})
  \,. $$
% in parctial implementations we suggest using softplus to numerically comptue the last
% term, since it is typically implemented in a floating-point stable fashion.
The numerically estimated derivative of the penalty term wrt $\log \alpha$ using forward
differences seems to be following \eqref{eq:real-kl-div-deriv-log} very closely (up to
sampling error) see fig.~\ref{fig:molchanov-derivative-replica}.

@@

@@

% TLDR: we use Wirtinger calculus!
Following \citep{trabelsi_deep_2017}, we naturally identify $\cplx$ with $\real^2$ via $
  z_\mathrm{r} + j z_\mathrm{i} \mapsto (z_\mathrm{r}, z_\mathrm{i})
$, storing the real and imaginary parts in interleaved format, and emulate $\cplx$-algebra
using $\real$-valued arithmetic.

% essential intro into Wirtinger calculus (CR)
% https://math.stackexchange.com/a/444493
Wirtinger ($\cplx\real$) calculus relies on the natural identification of $\cplx$ with $
  \real \times \real
$, and regards $
  f\colon \cplx \to \cplx
$ as an equivalent in algebraic sense function $F\colon \real^2 \to \cplx$ defined $
  f(z) = f(u + jv) = F(u, v)
$. Under this identification, we define two formal derivative operators
$
  \tfrac{\partial}{\partial z}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      - j \tfrac{\partial}{\partial v}
    \bigr)
$ and $
  \tfrac{\partial}{\partial \conj{z}}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      + j \tfrac{\partial}{\partial v}
    \bigr)
$ and the differential of $f$ at $z \in \cplx$ as
$$
df(z) \big\vert_{z=u + jv}
  = \frac{\partial f}{\partial z} dz
    + \frac{\partial f}{\partial \conj{z}} d\conj{z}
   % = \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    - j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % + \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    + j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % = \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du - j \frac{\partial F}{\partial v} du
   %    + j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   % + \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du + j \frac{\partial F}{\partial v} du
   %    - j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   = \frac{\partial F}{\partial u} du
     + \frac{\partial F}{\partial v} dv
   = dF(u, v)
  \,, $$
where $dz = du + j dv$ and $d\conj{z} = du - j dv$. This implies that the complex argument
and its conjugate are treated as independent variables. Cauchy-Riemann conditions $
  -j \tfrac{\partial F}{\partial v} = \tfrac{\partial F}{\partial u}
$ impose a rigid structure on real and imaginary parts of $F$ and can be expressed as $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$ in this notation. Thus $\cplx\real$ calculus subsumes the usual $\cplx$-calculus on
holomorphic functions, when $f(z)$, regarded as $f(z, \conj{z})$, is constant with respect
to $\conj{z}$. The usual rules of calculus, like chain and product rules, follow directly
from the definition of the operators, e.g.
$$
  \frac{\partial (f\circ g)}{\partial z}
    = \frac{\partial f(g(z))}{\partial g} \frac{\partial g(z)}{\partial z}
    + \frac{\partial f(g(z))}{\partial \conj{g}} \frac{\partial \conj{g(z)}}{\partial z}
    % = \nabla G \nabla F
  \,. $$
% Show that $dh(z) = df(g(z)) dg(z)$.
% $h(z) = f(g(z)) = F(G_r(u, v), G_i(u, v)) = H(u, v)$
% $$
% dH
%   = \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial u} du
%   + \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial v} dv
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial u} du
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial v} dv
%   = \tfrac{\partial F}{\partial r} d\Re g(z)
%   + \tfrac{\partial F}{\partial i} d\Im g(z)
%   = df(\Re g(z) + j\Im g(z)) d g(z)
%   = df(g(z)) d g(z)
%   \,. $$

In machine learning tasks the target objective is typically empirical risk, which has to
be real-valued to be minimized. Nevertheless, the expression of the $\cplx\real$ gradient
is compatible with what is expected, when $f$ is treated like a $\real^2$ function. For a
real-valued $f\colon \cplx \to \real$ we have $\conj{f} = f$, which implies $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial \conj{f}}{\partial \conj{z}}
    = \conj{\tfrac{\partial f}{\partial z}}
$, whence
$$
df
  = \tfrac{\partial f}{\partial z} dz
    + \tfrac{\partial f}{\partial \conj{z}} d\conj{z}
  % = \conj{\tfrac{\partial \conj{f}}{\partial \conj{z}}} dz
  % + \conj{\conj{\tfrac{\partial f}{\partial \conj{z}}} dz}
  = 2 \Re \Bigl(
    \conj{\tfrac{\partial f}{\partial \conj{z}}} dz
  \Bigr)
  = 2 \Re \bigl(
    \tfrac{\partial f}{\partial z} dz
  \bigr)
  \,. $$
Thus, the gradient of $f$ at $z$ is given by $
  \nabla_z f(z)
    = \tfrac{\partial F}{\partial u}
      + j \tfrac{\partial F}{\partial v}
$. Therefore, Wirtinger calculus makes it possible to reuse $\real$-backpropagation.

>>>>

$C = 0$. The entropy of proper $\cplx$-Gaussians is

@@

    \tag{\ref{eq:c-gauss-entropy-derivation}'}\,.

@@

We consider two priors for Bayesian dropout: log-uniform and $\cplx$-Gaussian empirical
Bayes, \citep{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}.
For a fully factorized approximation $q(\omega)$ and factorized prior belief $\pi(\omega)$,
the divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q(\omega) \| \pi(\omega))
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = \sum_{ij} \bigl(
        - \mathbb{H}(q(\omega_{ij}))
        - \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
      \bigr)
    \,.
\end{equation}
We use approximation \eqref{eq:c-gauss-vi} and omit subscripts ${ij}$ below for brevity.

@@

\begin{equation}  \label{eq:nonchisq-density}

@@

In \eqref{eq:ard-kl-div-real} we need $
$:

@@

    % &= \int_0^\infty f_W(x) \log x dx

@@

    \biggr) \log{(x)} dx

@@

which satisfies $

@@

Differentiating the series within $g_m$ yields
\begin{equation}  \label{eq:g-m-deriv}
  \frac{d}{d x} g_m(x)
    % = e^{-x} \sum_{n\geq 1} \frac{x^{n-1}}{(n-1)!} \psi(n + \tfrac{m}2) - g_m(x)
    % = e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} \psi(n + \tfrac{m}2 + 1) - g_m(x)
    % = e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} (
    %     \psi(n + \tfrac{m}2 + 1) - \psi(n + \tfrac{m}2)
    % )
    % = e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} \tfrac1{n + \tfrac{m}2}
    = e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} (n + \tfrac{m}2)^{-1}
    \,.
\end{equation}
Since $
    e^t  = \sum_{n\geq 0} \tfrac{t^n}{n!}
$ on $\mathbb{R}$, for $\alpha\neq 0$ we have
% $
%   \int_0^x t^{\alpha-1} dt
%     = \tfrac{x^\alpha}{\alpha}
% $ and
\begin{equation}  \label{eq:g-m-deriv-int-1}
  \sum_{n\geq 0} \int_0^x \frac{t^{n+\alpha-1}}{n!} dt
    % = \int_0^x \sum_{n\geq 0} \frac{t^{n+\alpha-1}}{n!} dt
    = \int_0^x t^{\alpha-1} e^t dt
    \,,
  % the MCT ([0, x], \mathcal{B}([0, x]), dx)
\end{equation}
by the monotone convergence theorem, whence
\begin{equation}  \label{eq:g-m-deriv-int-2}
  \eqref{eq:g-m-deriv}
    = x^{-\tfrac{m}2} e^{-x} \sum_{n\geq 0}
        \frac{x^{n + \tfrac{m}2}}{n!} (n + \tfrac{m}2)^{-1}
    % = x^{-\tfrac{m}2} e^{-x} \sum_{n\geq 0}
    %     \frac1{n!} \int_0^x t^{n + \tfrac{m}2 - 1} dt
    % = x^{-\tfrac{m}2} e^{-x}
    %     \int_0^x t^{\tfrac{m}2 - 1} \sum_{n\geq 0} \frac{t^n}{n!} dt
    = x^{-\tfrac{m}2} e^{-x}
        \int_0^x t^{\tfrac{m}2 - 1} e^t dt
    \,.
\end{equation}
The substitution $u^2 = t$ on $[0, \infty]$ with $2u du = dt$ yields
\begin{equation}  \label{eq:g-m-deriv-int-3}
  \frac{d}{d x} g_m(x)
    % = x^{-\tfrac{m}2} e^{-x}
    %     \int_0^x t^{\tfrac{m}2 - 1} e^t dt
    = 2 x^{-\tfrac{m}2} e^{-x}
        \int_0^{\sqrt{x}} u^{m - 1} e^{u^2} du
    \,.
\end{equation}
In particular, for $m=1$ we get
\begin{equation}  \label{eq:g-m-series-one}
  g_1(x)
    = - \gamma - 2 \log{2}
    + e^{-x} \sum_{n\geq 0} \frac{x^n}{n!} \sum_{p=1}^n \frac2{2p - 1}
    \,,
\end{equation}
and the derivative \eqref{eq:g-m-deriv} can be computed thus
\begin{equation}  \label{eq:g-m-deriv-one}
  \frac{d}{d x} g_1(x)
    = 2 \tfrac{F(\sqrt{x})}{\sqrt{x}}
    \,,
\end{equation}
using $
  F \colon \mathbb{R} \to \mathbb{R}
  \colon x \mapsto e^{-x^2} \int_0^x e^{u^2} du
$ Dawson's integral (special function).

The derivative of \eqref{eq:ard-kl-div-real} with respect to $\alpha$ is
$$
\tfrac{d}{d \alpha} g_1(\tfrac1{2\alpha})
  = -\tfrac1{2 \alpha^2} g_1'(\tfrac1{2\alpha})
  % = -\tfrac1{2 \alpha^2} 2 \tfrac{F(\sqrt{\tfrac1{2\alpha}})}{\sqrt{\tfrac1{2\alpha}}}
  % = -\tfrac1{2 \alpha^2} 2 \tfrac{F(\tfrac1{\sqrt{2\alpha}})}{\tfrac1{\sqrt{2\alpha}}}
  = -\tfrac1{\alpha} \sqrt{\tfrac2{\alpha}} F(\tfrac1{\sqrt{2\alpha}})
  \,. $$
Since $\alpha$ is nonnegative, it is typically parameterized via its logarithm and computed
when needed. Thus the gradient of \eqref{eq:ard-kl-div-real} w.r.t $\log \alpha$ is
\begin{equation}  \label{eq:real-kl-div-deriv-log}
  \frac{d \eqref{eq:ard-kl-div-real}}{d\log \alpha}
  % \frac{d}{d\log \alpha}
  %   \frac12 \mathbb{E}_{z \sim \mathcal{N}(0,1)}
  %     \log \lvert z + \tfrac1{\sqrt{\alpha}} \rvert^2
    % = \frac12 \frac{d\alpha}{d\log \alpha}
    %   \frac{d}{d\alpha} \bigl(\mathbb{E}\cdots \bigr)
    % = - \frac\alpha2 \tfrac1{\alpha}
    %   \sqrt{\tfrac2{\alpha}} F(\tfrac1{\sqrt{2\alpha}})
    = - \frac1{\sqrt{2\alpha}} F\bigl(\tfrac1{\sqrt{2\alpha}}\bigr)
    \,.
\end{equation}

We replicated the simulation experiment in \citep{molchanov_variational_2017} and recovered
the same nonlinear approximation result reported therein. We sampled $10^7$ draws of  and 
The coefficients are within $1\%$ relative tolerance of the reported.
$$
  \eqref{eq:ard-kl-div-real}
% - KL(\mathcal{N}(w \mid \mu, \alpha \mu^2) \|
%     \tfrac1{\lvert w \rvert})
%   \approx
%     k_1 \sigma(k_2 + k_3 \log \alpha)
%     + C \big\vert_{C = -k1}
%     - \tfrac12 \log (1 + e^{-\log \alpha})
  % = \tfrac12 \log \alpha
  %   - \mathbb{E}_{\xi \sim \mathcal{N}(1, \alpha)}
  %   \log{\lvert \xi \rvert} + C
  \approx
    \frac12 \log{\bigl(1 + e^{-\log \alpha}\bigr)}
    + k_1 \sigma\bigl(- (k_2 + k_3 \log \alpha)\bigr)
    % - k_1 \bigl(1 - \sigma(k_2 + k_3 \log \alpha)\bigr)
  \,. $$
% in parctial implementations we suggest using softplus to numerically comptue the last
% term, since it is typically implemented in a floating-point stable fashion.

@@

The numerically estimated derivative of the penalty term wrt $\log \alpha$ using forward
differences seems to be following \eqref{eq:real-kl-div-deriv-log} very closely (up to
sampling error) see fig.~\ref{fig:molchanov-derivative-replica}.

@@

% essential intro into Wirtinger calculus (CR)
% https://math.stackexchange.com/a/444493
Wirtinger ($\cplx\real$) calculus relies on the natural identification of $\cplx$ with $
  \real \times \real
$, and regards $
  f\colon \cplx \to \cplx
$ as an equivalent in algebraic sense function $F\colon \real^2 \to \cplx$ defined $
  f(z) = f(u + jv) = F(u, v)
$. Within this framework the differential of $f$ at $z \in \cplx$ is
$$
df(z) \big\vert_{z=u + jv}
  = \frac{\partial f}{\partial z} dz
    + \frac{\partial f}{\partial \conj{z}} d\conj{z}
   % = \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    - j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % + \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    + j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % = \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du - j \frac{\partial F}{\partial v} du
   %    + j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   % + \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du + j \frac{\partial F}{\partial v} du
   %    - j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   = \frac{\partial F}{\partial u} du
     + \frac{\partial F}{\partial v} dv
   = dF(u, v)
  \,, $$
with formally defined derivatives $
  \tfrac{\partial}{\partial z}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      - j \tfrac{\partial}{\partial v}
    \bigr)
$ and $
  \tfrac{\partial}{\partial \conj{z}}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      + j \tfrac{\partial}{\partial v}
    \bigr)
$, and differentials $dz = du + j dv$ and $d\conj{z} = du - j dv$. This implies that the
complex argument and its conjugate are treated as independent variables. Cauchy-Riemann
conditions $
  -j \tfrac{\partial F}{\partial v} = \tfrac{\partial F}{\partial u}
$ scan be expressed as $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$ in this notation. Thus $\cplx\real$ calculus subsumes the usual $\cplx$-calculus on
holomorphic functions, when $f(z)$, regarded as $f(z, \conj{z})$, is constant with respect
to $\conj{z}$. The usual rules of calculus, like chain and product rules, follow directly
from the definition of the operators, e.g.
$$
  \frac{\partial (f\circ g)}{\partial z}
    = \frac{\partial f(g(z))}{\partial g} \frac{\partial g(z)}{\partial z}
    + \frac{\partial f(g(z))}{\partial \conj{g}} \frac{\partial \conj{g(z)}}{\partial z}
    % = \nabla G \nabla F
  \,. $$
% Show that $dh(z) = df(g(z)) dg(z)$.
% $h(z) = f(g(z)) = F(G_r(u, v), G_i(u, v)) = H(u, v)$
% $$
% dH
%   = \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial u} du
%   + \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial v} dv
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial u} du
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial v} dv
%   = \tfrac{\partial F}{\partial r} d\Re g(z)
%   + \tfrac{\partial F}{\partial i} d\Im g(z)
%   = df(\Re g(z) + j\Im g(z)) d g(z)
%   = df(g(z)) d g(z)
%   \,. $$

In machine learning tasks the target objective is typically empirical risk, which has to
be real-valued to be minimized. Nevertheless, the expression of the $\cplx\real$ gradient
is compatible with what is expected, when $f$ is treated like a $\real^2$ function. For a
real-valued $f\colon \cplx \to \real$ we have $\conj{f} = f$, which implies $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial \conj{f}}{\partial \conj{z}}
    = \conj{\tfrac{\partial f}{\partial z}}
$, whence
$$
df
  = \tfrac{\partial f}{\partial z} dz
    + \tfrac{\partial f}{\partial \conj{z}} d\conj{z}
  % = \conj{\tfrac{\partial \conj{f}}{\partial \conj{z}}} dz
  % + \conj{\conj{\tfrac{\partial f}{\partial \conj{z}}} dz}
  = 2 \Re \Bigl(
    \conj{\tfrac{\partial f}{\partial \conj{z}}} dz
  \Bigr)
  = 2 \Re \bigl(
    \tfrac{\partial f}{\partial z} dz
  \bigr)
  \,. $$
Thus, the gradient of $f$ at $z$ is given by $
  \nabla_z f(z)
    = \tfrac{\partial F}{\partial u}
      + j \tfrac{\partial F}{\partial v}
$.

Natural identification of $\cplx$ with $\real\times \real$, storing the real and imaginary
parts in interleaved format, emulation of $\cplx$-algebra using $\real$-valued arithmetic,
and Wirtinger calculus make it possible to reuse $\real$-backpropagation and existing
auto-differentiation frameworks, \citep{trabelsi_deep_2017}.
